---
title: Health and economic impact analysys report of storms and other severe weather
  events, based on NOAA storm database
author: "Nuno Melo"
date: "August 18, 2015"
output:
  html_document:
    keep_md: yes
    number_sections: yes
    toc: yes
  pdf_document: default
---

# Synopsis  
**describes and summarizes your analysis in at most 10 complete sentences**  
In this report we aim to describe the economic and health impact caused by storms and other severe weather events, based on NOAA storm database. In particular it answers two questions:  

1. which types of events are most harmful to population health?
2. which types of events have the greatest economic consequences?

# Data description
The data for this analysys was downloaded from the Coursera web [site](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2in) [47Mb]. It is a comma-separated-value file compressed via the bzip2 algorithm to reduce its size.

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete. There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined. See below:  

* [National Weather Service Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

*  [National Climatic Data Center Storm Events FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

# Data Processing  
We first load our required R packages

## Loading required packages
```{r load packages}
library(dplyr)
library(lubridate)
library(ggplot2)
library(ggvis)
library(xtable)
```

## R Session Software environment used in the analysis
This is the information about R, the OS and attached or loaded packages
```{r SW Environment}
sessionInfo()
```

## Downloading data
The name of the file containing the data is *StormData.csv.bz2* and will be downloaded directly, from the Coursera web [site](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2in) [47Mb], to your working directory, in case the file does not already exists


```{r download data}
if (!file.exists("StormData.csv.bz2"))
        download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
                      "StormData.csv.bz2", quiet = FALSE, mode = "wb")

```

## Reading the *StormData.csv.bz2* in csv format into R
```{r read data, cache=TRUE}

data <- read.csv("StormData.csv.bz2")
```

## Checking the data 
After reading the data we check the raw data summary info and the first few rows 
```{r checking the data, cache = TRUE}
glimpse(data)
head(data) 

```

## Processing data

```{r processing data}

# From the dataset we select first only the variables relevant to our analysis, and create a new dataset storm_data.
# Pls observe we keep the original data intact throughout our analysys and calculations. Transformations are done in storm_data

selected_variables <- c("EVTYPE", "FATALITIES", "INJURIES", 
                        "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP") 
storm_data <- tbl_df(data) %>% select(one_of(selected_variables))

# Normalizing and calculating the economic impact

# We first normalize the expenditures as per described in 
# the National Weather Service Storm # Data Documentation.
# Normalized values will be stored in PROPDMG_NORM and CROPDMG_NORM 
# according to the following rules:
# B is billiion, i.e. 1000000000
# M is millions, i.e. 1000000
# K is thousand, i.e. 1000
# All other levels will not be taken in consideration, 
# because there is not enough information, and we will set them as NA

# Auxliary function to nomarlize PROPDMGEXP and CROPDMGEXP
ExpenditureNormalization <- function(exp) {
        new_exp <- 0
        
        new_exp[grep("K|k", exp)] <- "1000"
        new_exp[grep("M|m", exp)] <- "1000000"
        new_exp[grep("B|b", exp)] <- "1000000000"
        as.numeric(new_exp)
}

# Creating two new variables PROPDMG_NORM and CROPDMG_NORM with
# the total damage expenditures (values in $)
storm_data <- storm_data %>% mutate(
        PROPDMG_NORM = PROPDMG * ExpenditureNormalization(PROPDMGEXP),
        CROPDMG_NORM = CROPDMG * ExpenditureNormalization(CROPDMGEXP))

# Creating a summarized dataset with the sum of fatalities,
# injuries and fatalities + injuries per event type EVTYPE

summary_calculations <- storm_data %>% group_by(EVTYPE) %>%
        summarise(
                fatalities = sum(FATALITIES, na.rm = TRUE),
                injuries= sum(INJURIES, na.rm = TRUE),
                total_human = fatalities + injuries,
                property = sum(PROPDMG_NORM, na.rm = TRUE),
                crop = sum(CROPDMG_NORM, na.rm = TRUE),
                total_economic = property + crop
        )


```

# Graphs

```{r graphs}
df_graph_human <- summary_calculations %>% top_n(10, total_human)
graph_human <- ggplot(df_graph_human, aes(EVTYPE)) +
        geom_bar(aes(weight=total_human)) +
        labs(title = "Top 10 events most harmful to population health",
             x = "Event", y = "Human fatalities plus injuries") +
        theme(axis.text.x = element_text(angle = 90, hjust = 1))

df_graph_economic <- summary_calculations %>% top_n(10, total_economic)
graph_economic <- ggplot(df_graph_economic, aes(EVTYPE)) + geom_bar(aes(weight=total_economic)) +
        labs(title = "Top 10 events  with the greatest economic consequences\n
             Property plus Crops",
             x = "Event", y = "Dolars") +
        theme(axis.text.x = element_text(angle = 90, hjust = 1))

```     

# Results 

```{r results}

#  Selecting in descending order the "Top 10" events causing the 
#  highest fatalities, injuries, fatalities + injuries, property, 
#  crop and property + crop damages

top10_fatalities <- summary_calculations %>% top_n(10, fatalities) %>%
        select(EVTYPE, fatalities) %>% arrange(desc(fatalities))

top10_injuries <- summary_calculations %>% top_n(10, injuries) %>%
        select(EVTYPE, injuries) %>% arrange(desc(injuries))

top10_human <- summary_calculations %>% top_n(10, total_human) %>%
        select(EVTYPE, total_human) %>% arrange(desc(total_human))

top10_property <- summary_calculations %>% top_n(10, property) %>%
        select(EVTYPE, property) %>% arrange(desc(property))

top10_crop <- summary_calculations %>% top_n(10, crop) %>%
        select(EVTYPE, crop) %>% arrange(desc(crop))

top10_economic <- summary_calculations %>% top_n(10, total_economic) %>%
        select(EVTYPE, total_economic) %>% arrange(desc(total_economic))

```

## Population Health Impact
The events most harmful to population health are Tornados as shown per the **graph** and **table** below. They describe the **total fatalities plus injuries** for the 10 most impacting events.

```{r plot graph_human, echo=FALSE}

graph_human
top10_human

```

Below we present separately the the number of fatalities and injuries for the 10 most impacting events:    

### Fatalities
```{r plot top10_fatalities, echo=FALSE}
top10_fatalities
```

### Injuries

```{r plot top10_injuries, echo=FALSE}
top10_injuries
```

## Economical consequences
The events most harmful to population health are Floods as shown per the **graph** and **table** below. They describe **total property plus crop** economical impact for the 10 most impacting events.

```{r plot graph_economic, echo=FALSE}
graph_economic
top10_economic
```

Below we present in detailing the property and crop impact in dolars for the 10 most impacting events:  

### Property

```{r plot top10_property, echo=FALSE}
top10_property
```

### Crop


```{r plot top10_crop, echo=FALSE}
top10_crop
```
